Categories of Optics

### 2018-06-22 paper.tex
Some thoughts on effectful optics

@@

\todo{Contributions:
  \begin{itemize}
  \item A definition of optics general enough to encompass lenses, setters, traversals, and others
  \item A definition of lawfulness that specialises in the correct way in all examples and allows us to easily derive ad-hoc laws for new kids of optic
  \item A universal property satisfied by the category of optics in the monoidal case
  \item A satisfying categorical description of lens laws as the homomorphism condition on a morphism between certain comonoids
  \item An attempt to isolate what $\Set$-like properties are being used in various places
  \end{itemize}
}

@@

Two optics can be composed as follows: \todo{I give possibly too much detail here.} We wish to construct a map
\begin{align*}
  &\left(\int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S')\right) \times \left(\int^{M \in \C} \C(R, M \otimes S) \times \C(M \otimes S', R')\right) \\ &
                                                                                                                                                                       \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
The product in $\Set$ preserves colimits, so in particular coends. Using this fact and the Fubini theorem for coends, the domain is isomorphic to
\begin{align*}
  \int^{(M, N) \in \C \times \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R').
\end{align*}
So by the universal property of coends, it suffices to construct natural maps
\begin{align*}
  & \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') \\ &
                                                                                                              \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
for each $M, N \in \C$. For this we use the composites
\begin{align*}
  &\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R')\\
  \to \,& \C(N \otimes S, N \otimes M \otimes A) \times \C(N \otimes M \otimes A', N \otimes S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') && \text{(functoriality of $N \otimes  -$)} \\
  \to \,& \C(R, N \otimes  M \otimes A) \times \C(N \otimes M \otimes A', R') && \text{(composition in $\C$)} \\
  \to \,&\int^{M \in \M} \C(R, M \otimes A) \times \C(M \otimes A', R') && \text{($\copr_{N \otimes M}$)}
\end{align*}
In other words, let $\rep{l'}{r'} : (R, R') \hto (S, S')$ and $\rep{l}{r} : (S, S') \hto (A, A')$ with $M$ the complement for $\rep{l'}{r'}$. The composite is then: $\rep{l}{r} \circ \rep{l'}{r'} = \rep{(M l)l'}{r'(Mr)}$.

@@

\begin{proposition}\label{prop-optic-is-cat}
  The above data form a category $\Optic_\otimes$.
\end{proposition}
\begin{proof}
  In~\cite[Section 6]{Doubles} this is proven abstractly, by exhibiting this category as the Kleisli category for a monad in the bicategory $\Prof$. Here we prove the properties directly.

  We check associativity of composition equationally. As in the definition of composition, the Fubini theorem for coends allows us to choose representatives for all three optics simultaneously. Suppose we have representatives of three optics $\rep{l_1}{r_1} : (R, R) \hto (S, S')$, $\rep{l_2}{r_2} : (S, S') \hto (A, A')$ and $\rep{l_3}{r_3} : (A, A') \hto (B, B')$, that have complements $M$, $N$ and $P$ respectively. Then:
  \begin{align*}
    (\rep{l_3}{r_3} \circ \rep{l_2, r_2}) \circ \rep{l_1}{r_1}
    &= \rep{(N \otimes l_3)l_2}{r_2(N \otimes r_3)} \circ \rep{l_1}{r_1} \\
    &= \rep{(M \otimes ((N \otimes l_3)l_2))l_1}{r_1(M \otimes (r_2(N \otimes r_3)))} \\
    &= \rep{(M \otimes N \otimes l_3)(M \otimes l_2)l_1}{r_1(M \otimes r_2)(M \otimes N \otimes r_3)} \\
    &= \rep{l_3}{r_3} \circ (\rep{(M \otimes l_2)l_1}{r_1(M \otimes r_2)}) \\
    &= \rep{l_3}{r_3} \circ (\rep{l_2}{r_2} \circ \rep{l_1}{r_1})
  \end{align*}

  For the unit laws, suppose we have $\rep{l}{r} : (S, S') \hto (A, A')$ with representative $M$. We calculate:
  \begin{align*}
    \id_{A, A'} \circ \rep{l}{r}
    &= \rep{\lambda^{-1}_A}{\lambda_{A'}} \circ \rep{l}{r} \\
    &= \rep{(M \otimes \lambda^{-1}_A) l}{r (M\otimes  \lambda_{A'}} \\
    &= \rep{(\lambda^{-1}_M \otimes  A) l}{r (\lambda_M \otimes A'} \\
    &= \rep{l}{r (\lambda_M \otimes A') (\lambda^{-1}_M \otimes A')} \\
    &= \rep{l}{r} \\
    \rep{l}{r} \circ \id_{S, S'}
    &= \rep{l}{r} \circ \rep{\lambda^{-1}_S}{\lambda_{S'}}  \\
    &= \rep{(I \otimes l)\lambda^{-1}_S}{\lambda_{S'} (I \otimes r)} \\
    &= \rep{(\lambda^{-1}_M \otimes S)l}{r (\lambda_{M} \otimes S')} \\
    &= \rep{l}{r (\lambda_{M} \otimes S')(\lambda^{-1}_M \otimes S')} \\
    &= \rep{l}{r}
  \end{align*}
  Where in both cases we have used the naturality of $\lambda$ and coend relations to move $\lambda^{-1}_M$ to the right-hand side of the coend, where it may be cancelled.
\end{proof}

@@

\begin{definition}
  Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a monoidal functor $a : \M \to [\C, \C]$. The action of an object $M \in \M$ on $A \in \C$ is written $MA$.
\end{definition}

@@

The main reason is that it corresponds precisely to the ad-hoc lens laws with no additional requirements, whereas finding a complement for a lens requires further conditions on the category $\C$.

@@

Ideally, we then also find a simplified description $\conctwice(S, A)$ for the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S)\] appearing in the definition of a lawful optic. We can then ``read off'' what conditions on are needed on a concrete optic to ensure that the corresponding element of $\Optic_\M(S, A)$ is lawful. With no offence intended, we will call these more conditions the `ad-hoc' laws.

@@

\begin{proposition}\label{prop-OpticImpliesLensLaws}
  A lens $p : S \hto A$ is a lawful lens in our sense, i.e.\ that $\outside(p) = \id_S$ and $\once(p) = \twice(p)$, if and only if the corresponding $\fput$ and $\fget$ for this lens obey the three lens laws.
\end{proposition}
\begin{proof}
  We begin by giving
  \[ \int^{M_1, M_2 \in \C} \C(S, M_1 \times A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S)\]
  the same treatment as $\Optic_\times(S, A)$. By the same sorts of arguments as above \todo{is this worth writing out?}, this is isomorphic to
  \[\conctwice_\times(S, A) := \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)\]
  with the isomorphism given by
  \begin{align*}
    \Phi(\repthree{l}{c}{r }) = (&\pi_2 l, \\
                                   &\pi_2 c (\pi_1 l \times A), \\
                                   &r(\pi_1 c (\pi_1 l \times A) \times A))
  \end{align*}
  Now suppose we are given a lens $(\fget, \fput)$, which in coend form is $\rep{[\id_S, \fget]}{\fput}$. Evaluating $\outside$ on this gives:
  \begin{align*}
    \outside(\rep{[\id_S, \fget]}{\fput}) = \fput [\id_S, \fget]
  \end{align*}
  so requiring $\outside(p) = \id_S$ is precisely the $\fget\fput$ law.

@@

If we ask more of our category $\C$, we can upgrade this result to show that a lens $S \hto A$ implies the existence of a complement $\C$ with $S \cong C \times A$. This doesn't appear to follow purely from the ad-hoc lens laws---an argument using a definition of lawfulness based on constant complements.

\begin{proposition}[{Generalisation of \cite[Corollary 13]{AlgebrasAndUpdateStrategies}}]
  Suppose $\C$ has pullbacks and that there is a morphism $x : 1 \to A$. If $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
  Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
  \[
    \begin{tikzcd}
      S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
      & C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
      & 1\ar[r, "x", swap] & A
    \end{tikzcd}
  \]
  which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
  \begin{align*}
    \fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
                                &= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
                                &= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
                                &= \id_S && \text{(by $\fget\fput$)}
                                            \intertext{and}
                                            [j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
                                &= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
                                &= [j\fput (i \times A), \pi_2] && \\
                                &= [jij\fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
                                &= [jij \pi_1 (i \times A), \pi_2] && \\
                                &= [jiji \pi_1, \pi_2] && \\
                                &= [\pi_1, \pi_2] && \\
                                &= \id_{C \times A}
  \end{align*}

@@

The laws for prisms are the obvious duals to the lens laws:
\begin{align*}
  \fmatching \; \freview &= \inr \\
  [\id_S, \freview] \fmatching &= \id_S \\
  (\fmatching \sqcup A) \fmatching &= \mathrm{in}_{1,3} \, \fmatching
\end{align*}

@@

The following is then exactly the dual of Proposition~\ref{prop-OpticImpliesLensLaws}.
\begin{proposition}\label{prop-OpticImpliesPrismLaws}
  If $p : S \hto A$ is a prism then the associated $\fmatching$ and $\freview$ functions satisfy the three prism laws. \qed
\end{proposition}

@@

\begin{proposition}
  An iso $\rep{l}{r }$ is lawful iff (as expected) $l$ and $r$ are mutual inverses.
\end{proposition}
\begin{proof}
  \[ \int^{M_1, M_2 \in \M} \C(S, M_1 \otimes A) \times \C(M_1 \otimes A, M_2 \otimes A) \times \C(M_2 \otimes A, S) \]
  specialises in this case to just
  \[ \C(S, A) \times \C(A, A) \times \C(A, S) \]
  The condition $\outside(\rep{l}{r }) = \id_S$ is the claim that $rl = \id_S$, and $\once(\rep{l}{r }) = (l, \id_A, r)$ is equal to $\twice(\rep{l}{r }) = (l, lr, r)$ iff $lr = \id_A$.
\end{proof}

\subsection{Setters}

@@

\begin{definition}[\cite{StrongFunctors}]
  Suppose $(\C, \otimes, I)$ is a monoidal category.  A \emph{(left) strong functor} is a functor $F : \C \to \C$ equipped with a natural transformation
  \begin{align*}
    \theta_{A,B} : A \otimes F B \to F(A \otimes B)
  \end{align*}
  such that both
  \[
    \begin{tikzcd}
      1 \otimes F A \ar[r, "\theta_{1,A}"] \ar[d, "\cong" left]  & F(1 \otimes A) \ar[d, "\cong" right] \\
      F A \ar[r, equals] & F A
    \end{tikzcd}
  \]
  and
  \[
    \begin{tikzcd}
      (A \otimes B) \otimes F C \ar[rr, "\theta_{A \otimes B, C}"] \ar[d, "\alpha_{A,B,FC}" left]  && F((A \otimes B) \otimes C) \ar[d, "F\alpha_{A,B,FC}" right] \\
      A \otimes (B \otimes F C) \ar[r, "A \otimes \theta_{B,C}" below] & A \otimes F(B \otimes C) \ar[r, "\theta_{A, B\otimes C}" below] & F(A \otimes (B \otimes C))
    \end{tikzcd}
  \]
  commute. A \emph{strong natural transformation} $\tau : (F,\theta) \to (G,\theta')$ is a natural transformation that respects the strength:
  \[
    \begin{tikzcd}
      A \otimes F B \ar[r, "\theta_{A,B}"] \ar[d, "A \otimes \tau_B" left]  & F(A \otimes B) \ar[d, "\tau_{A \otimes B}" right] \\
      A \otimes G B \ar[r, "\theta'_{A, B}"]& G(A \otimes B)
    \end{tikzcd}
  \]

@@

\begin{proposition}
A setter $p : S \hto A$ is lawful iff 
\begin{align*}
\fover \; \nameof{\id_A} &= \nameof{\id_S} \\
({-} \circ =)(\fover \times \fover) &= \fover ({-} \circ =)
\end{align*}
where  $\circ$ denotes internal composition and $\nameof{\id_A} : 1 \to \homC(A, A)$ is the `name of' $\id_A$, and similarly for $\id_S$.
\end{proposition}
\begin{proof}
We have a chain of isomorphisms
\begin{align*}
&\int^{M_1, M_2 \in \Strong_\C} \C(S, M_1 A) \times \C(M_1 A, M_2  A) \times \C(M_2 A, S) \\
&\cong \int^{M_1, M_2 \in \Strong_\C} \Strong_\C(S \times \homC(A, -), M_1) \times \Strong_\C(M_1 A \times \homC(A, -), M_2) \times \C(M_2 A, S) \\
&\cong \int^{M_2 \in \Strong_\C} \Strong_\C(S \times \homC(A, A) \times \homC(A, -), M_2) \times \C(M_2 A, S) \\
&\cong \C(S \times \homC(A, A) \times \homC(A, A), S) \\
&\cong \C(\homC(A, A) \times \homC(A, A), \homC(S,S))
\end{align*}
\todo{Worth writing out the rest?}
\end{proof}

@@

Experts will recognise $UR^*_{-, -}$ as the free applicative functor \cite{FreeApplicativeFunctors} on the functor $R_{X,Y} Z = X \times Y \to Z$.

@@

\subsection{Linear Lenses}
\newcommand{\ev}{\mathsf{ev}}
\newcommand{\coev}{\mathsf{coev}}

@@

\begin{align*}
  \Optic_\otimes((S, S'), (A, A')) &= \int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \\
                                   &\cong \int^{M \in \C} \C(S, M \otimes A) \times \C(M, \homC(A',S')) \\
                                   &\cong \C(S, \homC(A',S') \otimes A) \\
\end{align*}
where $\homC(A', S')$ denotes the internal hom. If $\C$ is cartesian, this is of course isomorphic to the pair of maps discussed in the previous section.

For an optic, the corresponding element of $\conc_\otimes((S, S'), (A, A'))$ will be called $\funzip$. The above isomorphism sends $\funzip$ to the element $\rep{\funzip }{\ev_{A', S'}}$, where $\ev_{A', S'} : \homC(A',S') \otimes A' \to S'$ is the evaluation map. In the other direction, given $\rep{l}{r }$ we have $\funzip = (\tilde r \otimes A)l$ where $\tilde r : M \to \homC(A', S')$ denotes the transpose of $r$.

@@

\subsection{Polymorphic Optics}
Haskell's optics \todo{find an older reference from DB theory?} allow \emph{polymorphic updates}, where the type of the codomain of the lens can be changed by an update, causing a corresponding change in the type of the domain. As an example, the lens into the first entry of a tuple has type
\begin{minted}{haskell}
first :: Lens (a, b) (a', b) a a'
\end{minted}
then:
\begin{minted}{haskell}
set first (1, 5) "hello" == ("hello", 5)
\end{minted}
Note that the type has been changed from $\texttt{(Int, Int)}$ to $\texttt{(String, Int)}$.

@@

\subsection{Effectful Optics}
\todo{todo}
Many attempts~\cite{ReflectionsOnMonadicLenses} \todo{more refs} to define effectful lenses have wrapped the result of one or both of $\fget$ and $\fput$ with a monad. There are disadvantages to this approach: it is not obvious what the laws ought to be and there is no clear generalisation to other varieties of optic. The definition of optic given in Section~\ref{sec:optics} suggests we instead set $\C$ to be the Kleisli category of some effectful monad.

>>>>

\newcommand{\mget}{\textsc{MGet}}
\newcommand{\mput}{\textsc{MPut}}
\newcommand{\munzip}{\textsc{Munzip}}

@@

\todo{Contributions:
  \begin{itemize}
  \item A definition of optics general enough to encompass lenses, setters, traversals, and others
  \item A definition of lawfulness that specialises in the correct way in all examples and allows us to easily derive concrete laws for new kids of optic
  \item A universal property satisfied by the category of optics in the monoidal case
  \item A satisfying categorical description of lens laws as the homomorphism condition on a morphism between certain comonoids
  \item An attempt to isolate what $\Set$-like properties are being used in various places
  \end{itemize}
}

@@

Two optics can be composed as follows: \todo{I give possibly too much detail here.} We wish to construct a map
\begin{align*}
  &\left(\int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S')\right) \times \left(\int^{M \in \C} \C(R, M \otimes S) \times \C(M \otimes S', R')\right) \\ &
                                                                                                                                                                       \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
The product in $\Set$ preserves colimits, so in particular coends. Using this fact and the Fubini theorem for coends, the domain is isomorphic to
\begin{align*}
  \int^{(M, N) \in \C \times \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R').
\end{align*}
So by the universal property of coends, it suffices to construct natural maps
\begin{align*}
  & \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') \\ &
                                                                                                              \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
for each $M, N \in \C$. For this we use the composites
\begin{align*}
  &\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R')\\
  \to \,& \C(N \otimes S, N \otimes M \otimes A) \times \C(N \otimes M \otimes A', N \otimes S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') && \text{(functoriality of $N \otimes  -$)} \\
  \to \,& \C(R, N \otimes  M \otimes A) \times \C(N \otimes M \otimes A', R') && \text{(composition in $\C$)} \\
  \to \,&\int^{M \in \M} \C(R, M \otimes A) \times \C(M \otimes A', R') && \text{($\copr_{N \otimes M}$)}
\end{align*}
In other words, let $\rep{l'}{r'} : (R, R') \hto (S, S')$ and $\rep{l}{r} : (S, S') \hto (A, A')$ with $M$ the complement for $\rep{l'}{r'}$. The composite is then: $\rep{l}{r} \circ \rep{l'}{r'} = \rep{(M \otimes l)l'}{r'(M \otimes r)}$.

@@

\begin{proposition}\label{prop-optic-is-cat}
  The above data form a category $\Optic_\otimes$.
\end{proposition}
\begin{proof}
  In~\cite[Section 6]{Doubles} this is proven abstractly, by exhibiting this category as the Kleisli category for a monad in the bicategory $\Prof$. Here we prove the conditions directly.

  We check associativity of composition equationally.  Suppose we have representatives of three optics $\rep{l_1}{r_1} : (R, R) \hto (S, S')$, $\rep{l_2}{r_2} : (S, S') \hto (A, A')$ and $\rep{l_3}{r_3} : (A, A') \hto (B, B')$, that have complements $M$, $N$ and $P$ respectively. As in the definition of composition, the Fubini theorem for coends allows us to choose these representatives simultaneously. Then:
  \begin{align*}
    (\rep{l_3}{r_3} \circ \rep{l_2}{r_2}) \circ \rep{l_1}{r_1}
    &= \rep{(N \otimes l_3)l_2}{r_2(N \otimes r_3)} \circ \rep{l_1}{r_1} \\
    &= \rep{(M \otimes ((N \otimes l_3)l_2))l_1}{r_1(M \otimes (r_2(N \otimes r_3)))} \\
    &= \rep{(M \otimes N \otimes l_3)(M \otimes l_2)l_1}{r_1(M \otimes r_2)(M \otimes N \otimes r_3)} \\
    &= \rep{l_3}{r_3} \circ (\rep{(M \otimes l_2)l_1}{r_1(M \otimes r_2)}) \\
    &= \rep{l_3}{r_3} \circ (\rep{l_2}{r_2} \circ \rep{l_1}{r_1})
  \end{align*}

  For the unit laws, suppose we have $\rep{l}{r} : (S, S') \hto (A, A')$ with representative $M$. We calculate:
  \begin{align*}
    \id_{A, A'} \circ \rep{l}{r}
    &= \rep{\lambda^{-1}_A}{\lambda_{A'}} \circ \rep{l}{r} \\
    &= \rep{(M \otimes \lambda^{-1}_A) l}{r (M\otimes  \lambda_{A'}} \\
    &= \rep{(\lambda^{-1}_M \otimes  A) l}{r (\lambda_M \otimes A'} \\
    &= \rep{l}{r (\lambda_M \otimes A') (\lambda^{-1}_M \otimes A')} \\
    &= \rep{l}{r} \\
    \rep{l}{r} \circ \id_{S, S'}
    &= \rep{l}{r} \circ \rep{\lambda^{-1}_S}{\lambda_{S'}}  \\
    &= \rep{(I \otimes l)\lambda^{-1}_S}{\lambda_{S'} (I \otimes r)} \\
    &= \rep{(\lambda^{-1}_M \otimes S)l}{r (\lambda_{M} \otimes S')} \\
    &= \rep{l}{r (\lambda_{M} \otimes S')(\lambda^{-1}_M \otimes S')} \\
    &= \rep{l}{r}
  \end{align*}
  In both cases we have used the naturality of $\lambda$ and the coend relations to move $\lambda^{-1}_M$ to the right-hand side, where it may be cancelled.
\end{proof}

@@

\begin{definition}
  Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a monoidal functor $a : \M \to [\C, \C]$. For two objects $M \in \M$ and $A \in \C$, the action $a(M)(A)$ is abbreviated $MA$.
\end{definition}

@@

The main reason is that it corresponds precisely to the concrete lens laws with no additional requirements, whereas finding a complement for a lens requires further conditions on the category $\C$.

@@

Ideally, we then also find a simplified description $\conctwice(S, A)$ for the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S)\] appearing in the definition of a lawful optic. We can then ``read off'' what conditions on are needed on a concrete optic to ensure that the corresponding element of $\Optic_\M(S, A)$ is lawful. We will call these conditions the \emph{concrete} laws.

It is worth emphasising that once a monoidal action has been chosen and a concrete description of the corresponding optics found, no further work is needed to show that the result forms a category of optics, with a subcategory of lawful optics. This is especially useful when devising new optic variants, as we do later.

@@

\begin{proposition}\label{prop-OpticImpliesLensLaws}
  A concrete lens given by $\fget$ and $fput$ is lawful (in our sense) if and it obeys the three lens laws.
\end{proposition}
\begin{proof}
  We begin by giving
  \[ \int^{M_1, M_2 \in \C} \C(S, M_1 \times A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S)\]
  the same treatment as $\Optic_\times(S, A)$. By the same sorts of arguments as above \todo{is this worth writing out?}, this is isomorphic to
  \[\conctwice_\times(S, A) := \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)\]
  with the isomorphism given by
  \begin{align*}
    \Phi(\repthree{l}{c}{r }) = (&\pi_2 l, \\
                                   &\pi_2 c (\pi_1 l \times A), \\
                                   &r(\pi_1 c (\pi_1 l \times A) \times A))
  \end{align*}
  Now suppose we are given a lens $(\fget, \fput)$, which in coend form is $\rep{[\id_S, \fget]}{\fput}$. Evaluating $\outside$ on this gives:
  \begin{align*}
    \outside(\rep{[\id_S, \fget]}{\fput}) = \fput [\id_S, \fget]
  \end{align*}
  so requiring $\outside(p) = \id_S$ is precisely the $\fget\fput$ law.

@@

If we ask more of our category $\C$, we can upgrade this result to show that a lens $S \hto A$ implies the existence of a complement $\C$ with $S \cong C \times A$. This doesn't appear to follow purely from the concrete lens laws---an argument that a definition of lawfulness based on constant complements is not the correct generalisation.

\begin{proposition}[{Generalisation of \cite[Corollary 13]{AlgebrasAndUpdateStrategies}}]
  Suppose $\C$ has pullbacks and that there is a morphism $x : 1 \to A$. If $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the concrete lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
  Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
  \[
    \begin{tikzcd}
      S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
      & C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
      & 1\ar[r, "x", swap] & A
    \end{tikzcd}
  \]
  which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
  \begin{align*}
    \fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
                                &= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
                                &= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
                                &= \id_S && \text{(by $\fget\fput$)}
                                            \intertext{and}
                                            [j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
                                &= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
                                &= [j\fput (i \times A), \pi_2] && \\
                                &= [jij\fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
                                &= [jij \pi_1 (i \times A), \pi_2] && \\
                                &= [jiji \pi_1, \pi_2] && \\
                                &= [\pi_1, \pi_2] && \\
                                &= \id_{C \times A}
  \end{align*}

@@

The concrete laws for prisms are the obvious duals to the lens laws:
\begin{align*}
  \fmatching \; \freview &= \inr \\
  [\id_S, \freview] \fmatching &= \id_S \\
  (\fmatching \sqcup A) \fmatching &= \mathrm{in}_{1,3} \, \fmatching
\end{align*}

@@

The following is then exactly the dual of Proposition~\ref{prop-OpticImpliesLensLaws}.
\begin{proposition}\label{prop-OpticImpliesPrismLaws}
  If $p : S \hto A$ is a prism then the associated $\fmatching$ and $\freview$ functions satisfy the concrete prism laws. \qed
\end{proposition}

@@

\begin{proposition}
  An iso $\rep{l}{r}$ is lawful iff (as expected) $l$ and $r$ are mutual inverses.
\end{proposition}
\begin{proof}
  \[ \int^{M_1, M_2 \in \M} \C(S, M_1 \otimes A) \times \C(M_1 \otimes A, M_2 \otimes A) \times \C(M_2 \otimes A, S) \]
  specialises in this case to just
  \[ \C(S, A) \times \C(A, A) \times \C(A, S) \]
  The condition $\outside(\rep{l}{r }) = \id_S$ is the claim that $rl = \id_S$, and $\once(\rep{l}{r }) = (l, \id_A, r)$ is equal to $\twice(\rep{l}{r }) = (l, lr, r)$ iff $lr = \id_A$.
\end{proof}

\subsection{Setters}\label{sec:setters}

@@

\begin{definition}[\cite{StrongFunctors}]\label{def:strong-functor}
  Suppose $(\C, \otimes, I)$ is a monoidal category.  A \emph{(left) strong functor} is a functor $F : \C \to \C$ equipped with a natural transformation
  \begin{align*}
    \theta_{A,B} : A \otimes F B \to F(A \otimes B)
  \end{align*}
  such that the strength commutes with the unitor:
  \[
    \begin{tikzcd}
      I \otimes F A \ar[r, "\theta_{1,A}"] \ar[d, "\cong" left]  & F(I \otimes A) \ar[d, "\cong" right] \\
      F A \ar[r, equals] & F A
    \end{tikzcd}
  \]
  and with associativity:
  \[
    \begin{tikzcd}
      (A \otimes B) \otimes F C \ar[rr, "\theta_{A \otimes B, C}"] \ar[d, "\alpha_{A,B,FC}" left]  && F((A \otimes B) \otimes C) \ar[d, "F\alpha_{A,B,FC}" right] \\
      A \otimes (B \otimes F C) \ar[r, "A \otimes \theta_{B,C}" below] & A \otimes F(B \otimes C) \ar[r, "\theta_{A, B\otimes C}" below] & F(A \otimes (B \otimes C))
    \end{tikzcd}
  \]
  A \emph{strong natural transformation} $\tau : (F,\theta) \to (G,\theta')$ is a natural transformation that respects the strength:
  \[
    \begin{tikzcd}
      A \otimes F B \ar[r, "\theta_{A,B}"] \ar[d, "A \otimes \tau_B" left]  & F(A \otimes B) \ar[d, "\tau_{A \otimes B}" right] \\
      A \otimes G B \ar[r, "\theta'_{A, B}"]& G(A \otimes B)
    \end{tikzcd}
  \]

@@

\begin{proposition}
A setter $p : S \hto A$ is lawful iff 
\begin{align*}
\fover \; \nameof{\id_A} &= \nameof{\id_S} \\
({-} \circ =)(\fover \times \fover) &= \fover ({-} \circ =)
\end{align*}
where  $\circ$ denotes internal composition and $\nameof{\id_X} : 1 \to \homC(X, X)$ denotes the `name of' $\id_X$.
\end{proposition}
\begin{proof}
We have a chain of isomorphisms
\begin{align*}
&\int^{M_1, M_2 \in \Strong_\C} \C(S, M_1 A) \times \C(M_1 A, M_2  A) \times \C(M_2 A, S) \\
&\cong \int^{M_1, M_2 \in \Strong_\C} \Strong_\C(S \times \homC(A, -), M_1) \times \Strong_\C(M_1 A \times \homC(A, -), M_2) \times \C(M_2 A, S) \\
&\cong \int^{M_2 \in \Strong_\C} \Strong_\C(S \times \homC(A, A) \times \homC(A, -), M_2) \times \C(M_2 A, S) \\
&\cong \C(S \times \homC(A, A) \times \homC(A, A), S) \\
&\cong \C(\homC(A, A) \times \homC(A, A), \homC(S,S))
\end{align*}
\todo{Worth writing out the rest?}
\end{proof}

@@

% Experts will recognise $UR^*_{-, -}$ as the free applicative functor \cite{FreeApplicativeFunctors} on the functor $R_{X,Y} Z = X \times Y \to Z$.

@@

\subsection{Linear Lenses}\label{sec:linear-lenses}
\newcommand{\ev}{\mathsf{ev}}
\newcommand{\coev}{\mathsf{coev}}

@@

\begin{align*}
  \Optic_\otimes((S, S'), (A, A')) &= \int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \\
                                   &\cong \int^{M \in \C} \C(S, M \otimes A) \times \C(M, \homC(A',S')) \\
                                   &\cong \C(S, \homC(A',S') \otimes A)
\end{align*}
where $\homC(A', S')$ denotes the internal hom. If $\C$ is cartesian, this is of course isomorphic to the pair of maps discussed in the previous section.

A concrete linear lens is therefore a map $\funzip : S \to \homC(A',S') \otimes A$. The above isomorphism sends $\funzip$ to the element $\rep{\funzip}{\ev_{A', S'}}$, where $\ev_{A', S'} : \homC(A',S') \otimes A' \to S'$ is the evaluation map. In the other direction, given $\rep{l}{r}$ we have $\funzip = (\tilde r \otimes A)l$ where $\tilde r : M \to \homC(A', S')$ denotes the transpose of $r$.

@@

\subsection{Polymorphic Optics}
Haskell's optics \todo{find an older reference from DB theory?} allow \emph{polymorphic updates}, where the type of the codomain of the lens can be changed by an update, causing a corresponding change in the type of the domain. As an example, the lens into the first entry of a tuple has type
\begin{minted}{haskell}
first :: Lens (a, b) (a', b) a a'
\end{minted}
then:
\begin{minted}{haskell}
set first (1, 5) "hello" == ("hello", 5)
\end{minted}
Note that the type has been changed from \mintinline{haskell}{(Int, Int)} to \mintinline{haskell}{(String, Int)}.

@@

\subsection{Effectful Optics}
\newcommand{\monact}{\ltimes}

Many proposed definitions of effectful lenses~\cite{ReflectionsOnMonadicLenses} have modified one or both of $\fget$ and $\fput$ to produce results wrapped in a monadic action. There are disadvantages to this approach: it is not obvious what the laws ought to be and there is no clear generalisation to other optic variants. The general definition of optic given in Section~\ref{sec:optics} suggests we instead work with the Kleisli category $\C_T$ of some effectful monad $T : \C \to \C$.

\todo{define Kleisli category?} For a morphism $f : X \to Y$ in $\C_T$, we write $\underline{f} : X \to TY$ for its underlying morphism in $\C$.
